---
title: "Data Analysis"
author: "Mingye"
date: "2024-12-03"
output: html_document
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(rvest)
library(janitor)
library(geosphere)
library(sf)
```

```{r}
# data import 
rat_df = read.csv("data/rat_2020_2024.csv")
food_scrap_df = 
  read.csv("data/food_scrap_drop_off.csv")
zhvi_df = read.csv("data/ny_zhvi_df_2020_2024.csv")
zori_df = read.csv("data/ny_zori_df_2020_2024.csv")
nyc_boroughs = st_read("Borough Boundaries.geojson")


head(rat_df)
head(food_scrap_df)
head(zhvi_df)
head(zori_df)

```

```{r}
# Filter for rows where the result indicates rodent activity
data_rodent <- rat_df %>% filter(grepl("Rat Activity", result, ignore.case = TRUE))

# Group by neighborhood (NTA), year, and month to count occurrences of rodent activity
rodent_activity_summary <- data_rodent %>% 
  group_by(nta, year, month) %>% 
  summarise(count = n(), .groups = 'drop') %>% 
  arrange(desc(count))

# Print the top neighborhoods with the highest rodent activity
print(head(rodent_activity_summary))
```

```{r}
# Filter data for the year 2020
rodent_activity_2020 <- rodent_activity_summary %>% filter(year == 2020)

# Create a bar plot for rodent activity in 2020
ggplot(data = rodent_activity_2020, aes(x = month, y = count)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Rodent Activity in 2020",
       x = "month",
       y = "Rodent Activity Count") 
```

Purpose:

The bar plot shows how rodent activity varies across months in 2020.
This helps identify seasonal trends, such as higher rodent activity in summer months.
Findings:

There is a clear peak in rodent activity during the summer months (July and August).
A significant drop in activity is observed in the spring (April).
Justification:

Understanding monthly trends in rodent activity can inform targeted pest control strategies during peak seasons.

```{r}
# Aggregate data by year and month
yearly_monthly_summary <- data_rodent %>%
  group_by(year, month) %>%
  summarise(total_activity = n(), .groups = 'drop') %>%
  arrange(year, month)

# Create a line plot showing rodent activity over time
ggplot(yearly_monthly_summary, aes(x = as.numeric(month), y = total_activity, color = as.factor(year), group = year)) +
  geom_line(size = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Monthly Rodent Activity by Year (2020-2024)",
       x = "Month",
       y = "Total Rodent Activity Reports",
       color = "Year") +
  theme_minimal() +
  theme(legend.position = "right")
```
Objective:
To identify patterns in rodent activity across months and years.
To explore temporal trends and assess year-over-year consistency.

Findings:
Rodent activity generally peaks during the summer months (June to September), regardless of the year.
In some years, there are anomalies (e.g., dips in October for 2024).
A gradual increase in activity is evident over the years, suggesting either more rodents or improved reporting mechanisms.

```{r}
# Analyze the distribution of food scrap drop-off sites by borough
borough_distribution <- food_scrap_df %>% 
  group_by(borough) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

# Display the borough distribution
print("Borough Distribution:")
print(borough_distribution)
```

```{r}
# Borough distribution plot
ggplot(borough_distribution, aes(x = reorder(borough, -count), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Distribution of Food Scrap Drop-off Sites by Borough",
       x = "Borough",
       y = "Number of Sites") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Objective:
Analyze the geographic distribution of food scrap drop-off sites across NYC boroughs.
Identify whether disparities exist in access to food scrap sites between boroughs.

Findings:
Manhattan has the highest number of drop-off sites, followed by Brooklyn.
Bronx, Queens, and Staten Island have significantly fewer sites.
The disparity suggests possible inequities in access to food scrap recycling programs, particularly for Staten Island residents.


```{r}
# Analyze the temporal patterns
# First, let's create a simplified season categorization
food_scrap_df <- food_scrap_df %>%
  mutate(
    operation_type = case_when(
      grepl("Year Round", open_month, ignore.case = TRUE) ~ "Year Round",
      grepl("Seasonal|Spring|Summer|Fall|Winter", open_month, ignore.case = TRUE) ~ "Seasonal",
      TRUE ~ "Specific Months"
    )
  )

# Create a summary of operational patterns
operation_summary <- food_scrap_df %>%
  group_by(operation_type) %>%
  summarise(count = n())

print("Operation Pattern Summary:")
print(operation_summary)

# Create a scatter plot of locations
ggplot(food_scrap_df, aes(x = longitude, y = latitude)) +
  geom_point(aes(color = borough), alpha = 0.6) +
  theme_minimal() +
  labs(title = "Geographic Distribution of Food Scrap Drop-off Sites",
       x = "Longitude",
       y = "Latitude") +
  theme(legend.position = "right")
```
Objective:
To analyze the spatial distribution of food scrap drop-off sites across boroughs.
To visualize clustering and disparities in drop-off site availability.

Findings:
Drop-off sites are densely clustered in Manhattan and Brooklyn.
Sparse distribution in Staten Island reflects fewer food scrap programs in this borough.
Queens and Bronx show moderate distribution but lack the density seen in Manhattan.
```{r}
plot_county <- ggplot(zhvi_df, aes(x = CountyName, fill = CountyName)) +
  geom_bar() +
  labs(
    title = "Demographic Differences by Borough",
    x = "Borough",
    y = "Count",
    fill = "Borough"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),  # Rotate and resize x-axis text
    plot.title = element_text(size = 16, face = "bold"),  # Enlarge title
    axis.title.x = element_text(size = 14),  # Enlarge x-axis label
    axis.title.y = element_text(size = 14),  # Enlarge y-axis label
    legend.position = "none"  # Remove legend to reduce clutter
  )

print(plot_county)

```
Objective:
Analyze the geographic distribution of food scrap drop-off sites across NYC boroughs.
Identify whether disparities exist in access to food scrap sites between boroughs.
Findings:
Manhattan has the highest number of drop-off sites, followed by Brooklyn.
Bronx, Queens, and Staten Island have significantly fewer sites.
The disparity suggests possible inequities in access to food scrap recycling programs, particularly for Staten Island residents.

```{r}
# Filter rodent activity data for July
rodent_july <- data_rodent %>% filter(month == 8)

# Assuming food_scrap_df is available, filter it for July operations
food_scrap_july <- food_scrap_df %>% filter(grepl("July", open_month, ignore.case = TRUE))

# Create a scatter plot showing both rodent activity and food scrap sites in July
ggplot() +
  geom_point(data = rodent_july, aes(x = longitude, y = latitude, color = "Rodent Activity"), alpha = 0.6) +
  geom_point(data = food_scrap_july, aes(x = longitude, y = latitude, color = "Food Scrap Sites"), alpha = 0.6) +
  theme_minimal() +
  labs(title = "Interaction of Rodent Activity and Food Scrap Sites in July",
       x = "Longitude",
       y = "Latitude",
       color = "Legend") +
  theme(legend.position = "right")

# Load necessary libraries
library(leaflet)

# Filter rodent activity data for July
rodent_july <- data_rodent %>% filter(month == 8)

# Define NYC boroughs data (assuming nyc_boroughs is preloaded)
# Colors for boroughs
colors <- setNames(c("#FF0005", "#F0FDF0", "#5030FF", "#FFEF00", "#DF00F0"), unique(nyc_boroughs$boro_name))

# Create the leaflet map
rat_map <- leaflet() %>%
  addProviderTiles(providers$OpenStreetMap, options = providerTileOptions(noWrap = TRUE)) %>%
  setView(lng = -74.0060, lat = 40.7128, zoom = 11) %>%
  addPolygons(data = nyc_boroughs,
              fillColor = ~colors[as.character(nyc_boroughs$boro_name)],
              fillOpacity = 0.5,
              color = colors,
              weight = 1,
              popup = ~nyc_boroughs$boro_name) %>%
  addCircleMarkers(data = rodent_july,
                   lng = ~longitude, lat = ~latitude,
                   radius = 3,
                   color = "#00F0FF",
                   fillOpacity = 0.5,
                   popup = paste("<strong>Borough: </strong>", rodent_july$borough,
                                 "<br><strong>Date: </strong>", format(as.Date(paste0(rodent_july$year, "-", rodent_july$month, "-01")), "%Y-%m")))

# Print the map
rat_map
```
Objective:
The map overlays rodent activity data and food scrap drop-off sites for July, a peak month for rodent activity. The visualization aims to understand the spatial relationship between these two factors.

Clustered Rodent Activity:

Rodent activity is heavily concentrated in high-density urban areas, particularly Manhattan and parts of Brooklyn.
Sparse activity is observed in Staten Island, which aligns with its lower population density.
Food Scrap Drop-off Sites:

The distribution of food scrap sites shows a notable overlap with high rodent activity areas, particularly in Manhattan and Brooklyn.
Staten Island has both fewer rodent activity reports and fewer food scrap sites, indicating potential inequity in access to waste management resources.





