---
---
title: "Regression Models: Explaining Rodent Activity"
output: html_document
---

author: "Yujing Fu, Hanrui Li"
date: "2024-11-30"
output: html_document
---

```{r message=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(rvest)
library(janitor)
library(modelr)
```


```{r}
# data import 
rat_df = read.csv("data/rat_2020_2024.csv")
food_scrap_df = 
  read.csv("data/food_scrap_drop_off.csv")
zhvi_df = read.csv("data/ny_zhvi_df_2020_2024.csv")
zori_df = read.csv("data/ny_zori_df_2020_2024.csv")
```

```{r}
# clean rat data
rat_tidydf = rat_df |> 
  mutate(inspection_date = substring(inspection_date, 1, 10)) |> 
  mutate(date = as.Date(inspection_date, format = "%m/%d/%Y")) |> 
  mutate(year = as.numeric(format(date, "%Y"))) |> 
  mutate(month = as.numeric(format(date, "%m"))) |> 
  mutate(day = as.numeric(format(date, "%d"))) |> 
  select(inspection_type, result, date, year, month, day, street_name, zip_code, borough,latitude, longitude, location)
```

```{r}
# clean zhvi and zori data
zhvi_tidydf =
  zhvi_df |> 
  janitor::clean_names() |> 
  mutate(region_name=as.numeric(region_name)) |> # Convert the data type to numeric
  rename(zip_code = region_name, county = county_name) |> # Rename 'region_name' to 'zip_code'
  mutate(county = gsub(" County", "", county)) |> 
  pivot_longer(
    cols = starts_with("x20"), 
    names_to = "date", 
    values_to = "house_price") |>  
  mutate(date = as.Date(gsub("^x", "", date), format = "%Y_%m_%d")) |> # Remove 'x' from date and convert to Date format
  mutate(year = as.numeric(format(date, "%Y"))) |> # Extract the year from the date
  mutate(month = as.numeric(format(date, "%m"))) |> # Extract the month from the date
  select(zip_code, county, date, year, month, house_price) |> # Select relevant columns
  drop_na() # Remove rows with missing values for data quality

# Clean and transform ZORI (rental price) dataset
zori_tidydf = 
  zori_df |> 
  janitor::clean_names() |> # Standardize column names to clearer version
  rename(zip_code = region_name, county = county_name) |> # Rename 'region_name' to 'zip_code'
  mutate(county = gsub(" County", "", county)) |> 
  pivot_longer(
    cols = starts_with("x20"), 
    names_to = "date", 
    values_to = "rental_price") |> 
  mutate(date = as.Date(gsub("^x", "", date), format = "%Y_%m_%d")) |> # Remove 'x' from date and convert to Date format
  mutate(year = as.numeric(format(date, "%Y"))) |> 
  mutate(month = as.numeric(format(date, "%m"))) |>
  select(zip_code, county, date, year, month, rental_price) |> # Select necessary columns
  drop_na() # Remove rows with missing values
```

```{r}
# zip code df
url = "https://p8105.com/data/zip_codes.html"
zip_code_html = read_html(url)

zip_code_df =
  zip_code_html |> 
  html_table() |> 
  as.data.frame() |> 
  janitor::clean_names() |> 
  mutate(borough = case_when(
    county == "Bronx" ~ "Bronx",
    county == "Kings" ~ "Brooklyn",
    county == "New York" ~ "Manhattan",
    county == "Queens" ~ "Queens",
    county == "Richmond" ~ "Staten Island",
    TRUE ~ "Other"
  )) |> 
  select(zip_code, county, neighborhood, borough) |> 
  drop_na()
```

```{r}
# merge the zori & zhvi with zip code based on both the `zip_code` and `county`
zori_merged = left_join(zori_tidydf, zip_code_df, by = c("zip_code", "county")) |> 
  select(-date)
zhvi_merged = left_join(zhvi_tidydf, zip_code_df, by = c("zip_code", "county")) |> 
  select(-date)
```


```{r}
# merge the zillow information
zillow_merged = merge(zori_merged, zhvi_merged, by = c("zip_code", "county", "borough", "neighborhood", "year", "month")) |> 
  drop_na()
```

```{r}
# define the binary variable of rodent activity
rodent_reg =
  rat_tidydf |> 
  mutate(rodent_bi = case_when(
    result == "Rat Activity" ~ 1,
    TRUE ~ 0
  )) 
```


```{r}
rodent_reg_merged = 
  left_join(zillow_merged, rodent_reg, by = c("zip_code", "borough", "year", "month"))
```


# Data Analysis

**rodent_bi = 1**: Indicates "Rat Activity" was found during the inspection. This means that there was a confirmed presence of rodent activity at the inspected location.

**rodent_bi = 0**: Indicates "No Rat Activity" or any other result different from "Rat Activity". This means there was no confirmed rodent activity at the inspected location.

## Is the rental price of houses significantly related to the rodent activity?

```{r}
# Logistic Regression: Rental Price vs Rodent Activity
model1 <- glm(rodent_bi ~ rental_price, family = binomial, data = rodent_reg_merged)

# Creating Regression Model Table
model1 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$log[\frac{P(\text{rodent activity} = 1)}{1 - P(\text{rodent activity} = 1)}] = -0.766 + 0.0000779 \times \text{rental price}$$

Where:

* $P(\text{rodent activity} = 1))$ is the probability of observing rodent activity.
* $\text{log}(\frac{P}{1-P})$ is the log-odds of rodent activity.

**Intercept ($\beta_0$):** -0.766

This represents the log-odds of rodent activity when the rental price is 0.

* Since the intercept value is negative ($\beta_0 = -0.766$), the log-odds of rodent activity are less than zero when the rental price is 0.
* Odds = $e^{\beta_0} = e^{-0.766} \approx 0.465$. This implies that, when the rental price is 0, the odds of rodent activity are approximately 0.465:1, meaning it's less likely than not.

**Coefficient for `rental_price` ($\beta_1$):** 0.0000779

The coefficient represents the change in the log-odds of rodent activity for a one-unit increase in the rental price.

* For every one-unit increase in the rental price, the log-odds of rodent activity increase by 0.0000779.
* Odds Ratio = $e^{\beta_1} = e^{0.0000779} \approx 1$. This indicates that a small increase in rental price slightly increases the odds of rodent activity.

**Statistical Significance:**

Both p-values for the intercept and coefficient are <2e-16 $\approx$ 0, meaning they are statistically significant at any common significance level (e.g., $\alpha = 0.05$). This suggests that both the intercept and the effect of rental price on rodent activity are statistically significant.

**Summary:** 

* Rental price has a statistically significant but very small positive relationship with rodent activity. While significant, the effect size is negligible, meaning rental price is not a strong predictor of rodent activity.
* This suggests that other factors likely play a more significant role in determining rodent activity in the dataset.

## Is the value of houses significantly related to the rodent activity?

```{r}
# Logistic Regression: House Value vs Rodent Activity
model2 = glm(rodent_bi ~ house_price, family = binomial, data = rodent_reg_merged)

# Creating Regression Model Table
model2 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$log[\frac{P(\text{rodent activity} = 1)}{1 - P(\text{rodent activity} = 1)}] = -0.451 + (-0.0000001) \times \text{house value}$$

**Intercept ($\beta_0$):** -0.451

This represents the log-odds of rodent activity when the house value is 0.

* Since `house_price` = 0 is unrealistic (as house prices cannot be zero), the intercept primarily serves as a baseline log-odds for the model.
* Odds = $e^{\beta_0} = e^{-0.451} \approx 0.637$. This means that the odds of rodent activity are approximately 0.637:1 when house price is 0 (hypothetically).

**Coefficient for `rental_price` ($\beta_1$):** -0.0000001

The coefficient indicates the change in log-odds of rodent activity for a one-unit increase in house price.

* For each one-unit increase in house price, the log-odds of rodent activity decrease by 0.0000001.
* Odds Ratio = $e^{\beta_1} = e^{-0.0000001} \approx 1$. This suggests that an increase in house price very slightly reduces the odds of rodent activity. However, the effect size is extremely small and likely negligible in practical terms.

**Statistical Significance:**

Both p-values for the intercept and coefficient $\approx$ 0, meaning they are statistically significant at any common significance level (e.g., $\alpha = 0.05$). This means the relationship between house price and rodent activity is statistically significant, but the effect size is so small that it has little practical importance.

**Summary:** 

* Higher house prices are slightly associated with lower rodent activity, but the effect is negligible.
* While statistically significant, the effect size is so small that house price is unlikely to be a meaningful predictor of rodent activity. Other variables are likely more important in explaining the variation in rodent activity.


## Is borough a confounder on the relationship between rodent activity and rental price? (Also between rodent activity and house value?)

### Is borough significantly related to the rental price?

```{r}
#Linear regression model: rental price and borough
model3 <- lm(rental_price ~ borough, data = rodent_reg_merged)

# Creating Regression Model Table
model3 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$\text{rental price} = 2100.03 + 805.85\times \text{Brooklyn} + 1148.49\times \text{Manhattan} + 379.75\times \text{Queens} + 45.98\times \text{Staten Island}$$
Where:

* $\beta_0$: Intercept (average rental price in the reference category, Bronx).
* $\beta_{1,2,3,4}$: Coefficients for the borough dummy variables. These represent the difference in average rental price between each borough and the Bronx.

**Intercept ($\beta_0$):** 2100.03

This represents the average rental price in the Bronx (the reference borough).

**Coefficient for borough ($\beta_{1,2,3,4}$):**

This indicates how much the average rental price in each borough (Brooklyn, Manhattan, Queens, Staten Island) is higher than in the Bronx. All are positive.

**Statistical Significance:**

The overall p-value for the model (p < 2e-16) indicates that borough is significantly related to rental price.
The p-values for each borough relative to the reference category (Bronx) are all statistically significant (p < 0.001), meaning rental prices differ significantly across boroughs.

**Summary:** 

Borough is significantly related to rental price, with notable differences in average rental prices across boroughs.

### Is borough significantly related to the rodent activity?

```{r}
#Regression model: rodent activity and borough
model4 <- glm(rodent_bi ~ borough, data = rodent_reg_merged)

# Creating Regression Model Table
model4 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$\text{rodent activity}=0.441+(−0.0847)\times \text{Brooklyn}+(−0.0350)\times \text{Manhattan}+(−0.1631)\text{Queens}+(−0.3443)\times \text{StatenIsland}$$
Where:

* $\beta_0$: Intercept (baseline probability of rodent activity in the reference borough, Bronx).
* $\beta_{1,2,3,4}$: Coefficients representing the difference in rodent activity between each borough and the Bronx.

**Intercept ($\beta_0$):** 0.441

This represents the baseline probability of rodent activity in the Bronx (the reference category).

**Coefficient for borough ($\beta_{1,2,3,4}$):**

The coefficient for each borough (Brooklyn, Manhattan, Queens, Staten Island) represents the difference in the average probability of rodent activity between each borough and the Bronx. For example, the probability of rodent activity in Brooklyn is 0.0847 lower than in the Bronx.

**Statistical Significance:**

All coefficients, including the intercept, have p < 0.001, indicating that the differences in rodent activity probabilities across boroughs are statistically significant.

**Summary:** 

Borough is significantly related to rodent activity.

### Regression model with both borough and rental price

```{r}
#Regression model testing for confounding
model5 <- glm(rodent_bi ~ borough + rental_price, data = rodent_reg_merged)

# Creating Regression Model Table
model5 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$\text{rodent acticvity} = 0.4179+(−0.0936)\times \text{Brooklyn}+(−0.0476)\times \text{Manhattan}+(−0.1673)\times \text{Queens}+(−0.3448)\times \text{Staten Island}+(0.000011)\times \text{rental price}$$

Where:

* $\beta_0:$ Intercept (baseline probability of rodent activity in the Bronx with rental price = 0).
* $\beta_{1,2,3,4}:$ Coefficients for boroughs (relative to Bronx).
* $\beta_5:$ Coefficient for rental price, indicating how rental price impacts rodent activity.

**Intercept ($\beta_0$):** 0.4179

The intercept represents the baseline probability of rodent activity in the Bronx when rental price is 0.

**Coefficient for borough ($\beta_{1,2,3,4}$):**

These coefficients represent the difference in probability of rodent activity between each borough and the Bronx (the reference borough).

**Coefficient for borough ($\beta_{5}$):** 0.000011

Rental price has a small positive effect on the probability of rodent activity.

**Statistical Significance:**

All predictors, including boroughs and rental price, are statistically significant (p < 0.001). This suggests that both borough and rental price significantly contribute to the prediction of rodent activity.

**Summary:** 

Borough remains a strong predictor of rodent activity. Rental price is also a significant predictor.

### Is borough significantly related to the house value?

```{r}
#Linear regression model: house value and borough
model6 <- lm(house_price ~ borough, data = rodent_reg_merged)

# Creating Regression Model Table
model6 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$\text{house value} = 434080.5+517358.1\times \text{Brooklyn}+759697.9\times \text{Manhattan}+303134.7\times \text{Queens}+216711.8\times \text{StatenIsland}$$

Where:

* $\beta_0$: Intercept (average rental price in the reference category, Bronx).
* $\beta_{1,2,3,4}$: Coefficients for the borough dummy variables, indicating differences in house value relative to the Bronx.

**Intercept ($\beta_0$):** 434080.5

The intercept represents the average house value in the Bronx (the reference category).

**Coefficient for borough ($\beta_{1,2,3,4}$):**

These coefficients indicate how much higher the average house value is in each borough compared to the Bronx. All positive.

**Statistical Significance:**

All borough coefficients ($p\approx 0$) are statistically significant, indicating significant differences in average house values between the Bronx and other boroughs. The overall model is significant, meaning that borough is a significant predictor of house value.

**Summary:** 

Borough is significantly related to house value, reflecting differences in property markets across New York City.

### Regression model with both borough and house value

```{r}
#Regression model testing for confounding
model7 <- glm(rodent_bi ~ borough + house_price, data = rodent_reg_merged)

# Creating Regression Model Table
model7 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$\text{rodent activity} = 0.4626+(−0.0590)\times \text{Brooklyn}+(0.0027)\times \text{Manhattan}+(−0.1480)\times \text{Queens}+(−0.3335)\times \text{Staten Island}+(−0.000000497)\times \text{house price}$$

Where:

* $\beta_0:$ Intercept, the baseline probability of rodent activity in the Bronx with house price = 0.
* $\beta_{1,2,3,4}:$ Coefficients for boroughs (compared to the Bronx).
* $\beta_5:$ Coefficient for house price, representing the effect of house price on rodent activity.

**Intercept ($\beta_0$):** 0.4626

The intercept represents the baseline probability of rodent activity in the Bronx (the reference category) when the house price is 0.

**Coefficient for borough ($\beta_{1,2,3,4}$):**

These coefficients represent the difference in the probability of rodent activity for each borough compared to the Bronx, controlling for house price.

**Coefficient for borough ($\beta_{5}$):** -0.000000497

House price has a very small negative effect on the probability of rodent activity.

**Statistical Significance:**

Manhattan is not statistically significant (p = 0.58), but all other borough coefficients are statistically significant (p < 0.001). The coefficient for house price is statistically significant, but the effect size is negligible.

**Summary:** 

Borough remains a significant predictor of rodent activity. House price also has a statistically significant effect.

### Regression model with borough, rental price, and house value

```{r}
#Regression model testing for confounding
model8 <- glm(rodent_bi ~ borough + rental_price + house_price, data = rodent_reg_merged)

# Creating Regression Model Table
model8 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

$$\text{rodent activity} = 0.3864+(−0.0714)\times \text{Brooklyn}+(−0.0137)\times \text{Manhattan}+(−0.1510)\times \text{Queens}+(−0.3252)\times \text{Staten Island}+(0.0000462)\times \text{rental price}+(−0.000000978)\times \text{house price}$$

Where:

* $\beta_0:$ Intercept, representing the baseline probability of rodent activity in the Bronx with rental price = 0 and house price = 0.
* $\beta_{1,2,3,4}:$ Coefficients for boroughs (relative to the Bronx).
* $\beta_5:$ Coefficient for rental price, representing the effect of rental price on rodent activity.
* $\beta_6:$ Coefficient for house price, representing the effect of house price on rodent activity.

**Intercept ($\beta_0$):** 0.3864

The intercept represents the baseline probability of rodent activity in the Bronx when rental_price=0 and house_price=0.

**Coefficient for borough ($\beta_{1,2,3,4}$):**

These coefficients represent the difference in the probability of rodent activity for each borough compared to the Bronx, controlling for rental price and house price.

**Coefficient for borough ($\beta_{5}$):** 0.0000462

Rental price has a very small positive effect on the probability of rodent activity.

**Coefficient for borough ($\beta_{6}$):** -0.000000978

House price has a very small negative effect on the probability of rodent activity.

**Statistical Significance:**

All borough coefficients are statistically significant (p<0.01). Rental price and house price are statistically significant (p=0), but their effects are very small.

**Summary:** 

Borough is a strong predictor of rodent activity. Rental price and house price are also statistically significant.

**In conclusion, Borough remains the dominant predictor of rodent activity in the model. Although rental price and house price are statistically significant, their effects are too small to have practical relevance. Here, we need to compare the models to decide which predictors to consider.**

### Comparison of models with or without rental price and house value

```{r}
cv1_df = 
  crossv_mc(rodent_reg_merged, 100)

cv1_df |> 
  mutate(
    mod_1  = map(train, \(df) glm(rodent_bi ~ borough, data = rodent_reg_merged)),
    mod_2  = map(train, \(df) glm(rodent_bi ~ borough + rental_price + house_price, 
                                  data = rodent_reg_merged))) |> 
  mutate(
    rmse_1 = map2_dbl(mod_1, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_2 = map2_dbl(mod_2, test, \(mod, df) rmse(model = mod, data = df))) |>
  select(starts_with("rmse")) |>
  pivot_longer(
    everything(), 
    names_to = "model", 
    values_to = "rmse", 
    names_prefix = "rmse_") |>
  ggplot(aes(x = model, y = rmse, group = model, fill = model)) + 
  geom_violin(alpha = 0.5) +
  theme_minimal()
```

The model excluding rental price and house value has a much higher root mean squared errors (RMSE) value than the model including rental price and house value. This means that the **model including rental price and house value** better fits our data.

## Testing for Interaction between rental price, house value, and borough

```{r}
#Regression model testing for confounding
model9 = glm(formula = rodent_bi ~ borough + rental_price + house_price + 
              borough*rental_price*house_price, data = rodent_reg_merged)

# Creating Regression Model Table
model9 |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(digits = 10)
```

There are significant interactions between borough and rental price as well as borough and house price, particularly in Brooklyn and Queens. This indicates that the relationship between rental price, house price, and rodent activity varies across different boroughs.

This model is considerably more complex, with numerous interaction terms. The results suggest that while borough remains a key factor in predicting rodent activity, the influence of rental and house prices varies by borough. However, many of the interaction effects are not statistically significant, implying that the added complexity may not be providing substantial predictive improvement.

Therefore, we need to check further.


## Comparison of models with and without interaction term:

```{r}
cv2_df = 
  crossv_mc(rodent_reg_merged, 100) 

cv2_df |> 
  mutate(
    mod_1  = map(train, \(df) glm(formula = rodent_bi ~ borough + rental_price + house_price 
                                  + borough*rental_price*house_price, data = rodent_reg_merged)),
    mod_2  = map(train, \(df) glm(formula = rodent_bi ~ borough + rental_price + house_price, 
                                  data = rodent_reg_merged))) |> 
  mutate(
    rmse_1 = map2_dbl(mod_1, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_2 = map2_dbl(mod_2, test, \(mod, df) rmse(model = mod, data = df))) |>
  select(starts_with("rmse")) |>
  pivot_longer(
    everything(), 
    names_to = "model", 
    values_to = "rmse", 
    names_prefix = "rmse_") |>
  ggplot(aes(x = model, y = rmse, group = model, fill = model)) + 
  geom_violin()
```

The model with an interaction term between borough, rental price, and home value has a lower RMSE value than the model without an interaction term. Ultimately, the model with interaction terms appears to be the best model.



